---
title: "Chapter 12 Notes"
subtitle: "mutate() and case_when()"
author: "Kylie George"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
---

# Quarto Notes

# Load Packages
```{r}
library("dslabs")
library("dplyr")
```


# Vectorized Conditionals
evaluate conditions element-by-element across a vector and return a new vector of the same length.

In other words, subsetting answers “What do I keep?” while vectorized conditionals answer “What value should I return for each element?”

```{r}
# 📖 Review
## Create vector x
x <- 1:6
## logical vector
x > 3

# 📖 Review
## logical subsetting
x[x > 3]
```
🗣 x[x > 3] subsets x to keep only values greater than 3.

```{r}
# 📖 Review
## replaces values
x[x > 3] <- 99
## ✅  print result
x
```
🗣 x[x > 3] subsets x to keep only the values greater than 3, and x[x > 3] <- 99 uses logical indexing to replace those matching values with 99.

## ifelse()
Base R describes ifelse() as returning a value with the same shape as the test, filled from yes or no.

📜 The SYNTAX: ifelse(test, yes, no)
test – logical vector (TRUE/FALSE/NA)
yes – value to insert where test is TRUE
no – value to insert where test is FALSE.
NA – if NA, the result is NA at that position.

🎯 Label the vector of numbers greater than 3 as “Large” and others as “Small”.

```{r}
x <- c(3, 2, 3, 6, 5)
length(x)
```
🗣 `x` is a vector with `{r} length(x)` values.

```{r}
ifelse(test = x >3, yes = "Large", no = "Small")
```
💡 Tip: Read this ifelse() statement like ‘If the vector is greater than 3, then return “Large”, otherwise return “Small”.’

🗣 The output positions for 3 and under are “Small” and numbers higher than 3 are “Large”.

🧐 Notice: Arguments, such as test, yes, and no in ifelse(), are often not required to be named inside of functions.


```{r}
length(ifelse(x > 3, "Large", "Small"))
```

🗣 The result has the same length as x. Nothing was dropped. Each position got a label.

## case_when()
dplyr describes case_when() as a general vectorized if-else that evaluates rules in order and uses the first match for each element. This makes it especially useful when you have several categories to assign.

📜 The SYNTAX: case_when(condition1 ~ value1, condition2 ~ value2, .default = default_value)

🎯 Label the vector of numbers greater than 3 as “Large” and others as “Small”.

```{r}
case_when(c(3, 2, 3, 6, 5) > 3 ~ "Large", .default = "Small")

```
🗣 The output positions for 3 and under are “Small” and numbers higher than 3 are “Large”.

📖 Why do we need both ifelse() and case_when()? case_when() is usually clearer once you have 3+ categories.

## More than Two Categories
ifelse() is designed for two categories. If you have more than two, you can nest ifelse() statements inside each other, but this quickly becomes hard to read. case_when() is designed for multiple categories and is much easier to read when you have more than two conditions.

```{r}
scores <- c(55, 65, 75, 85, 95)
```
🎯 Re-code scores less than 60 as “Low”, scores less than 80 as “Medium”, and all others as “High”.

### Nested ifelse()
🎯 Re-code scores less than 60 as “Low”, scores less than 80 as “Medium”, and all others as “High” using nested ifelse().

```{r}
ifelse(scores < 60, "Low",
       ifelse(scores < 80, "Medium", "High"))
```
🗣 This works, but the logic is nested inside the no argument of the first ifelse(), which becomes harder to read as the number of categories grows.

### Multiple Rule case_when()
case_when() avoids the ugly nesting that happens with repeated ifelse() calls. It is a general vectorized if-else: R checks the rules from top to bottom and uses the first rule that matches each element. That is why the order of the rules matters.

🎯 Code scores less than 60 as "Low", scores less than 80 as "Medium", and all others as "High" using case_when().

```{r}
case_when(scores < 60 ~ "Low",
          scores < 80 ~ "Medium",
          TRUE        ~ "High")
```
🗣 This reads more like a set of classification rules: first check for "Low", then check for "Medium", and assign "High" to everything else.

⚠️ The order of the rules matters because case_when() uses the first condition that matches.

🐱 This is a good place to be curious and test what happens when the rules are in the wrong order. The goal is not to break the code randomly; the goal is to see how case_when() makes its decision.
```{r}
case_when(
  scores < 80 ~ "Medium",
  scores < 60 ~ "Low",
  TRUE       ~ "High"
)
```
🗣 This does not work as intended. Any score less than 60 also satisfies scores < 80, so those values are labeled "Medium" before R ever checks the "Low" rule.

💡 Tip: write rules from most specific to more general, and end with a default rule such as TRUE ~ "High".

# mutate()

## Column Copy

### mutate() with Arithmetic

### case_when() inside mutate()

## Overwrite Column Values